What is an Ancestral Recombination Graph?

### 2022-03-15 paper.tex
Start on fig 6

\section*{ARGs and SPRs}
\citet{wiuf1999ancestry,wiuf1999recombination} recast
the coalescent with recombination
as a stocastic process along the genome rather than backwards in time.
In this formulation we begin at one end of a chromosome by
sampling a tree under the standard coalescent, and then reason
about how this tree is modified by recombination events
as we move along the genome. Although this algorithm is
significantly more complicated to simulate than Hudson's
backwards in time approach, it has led to substantial
insights into the process of recombination itself
and to the sequentially Markov Coalescent (SMC)
approximations~\citep{mcvean2005approximating,marjoram2006fast}
which are central to many modern methods.

@@

@@

\begin{figure}
\vspace{5em}
\caption{\label{fig-inferred-args}
ARGs inferred by different methods from a small dataset. Notes about these
ARGs.
}
\end{figure}

>>>>

\section*{ARGs and SPRs}
\citet{wiuf1999ancestry,wiuf1999recombination} recast
the coalescent with recombination
as a stochastic process along the genome rather than backwards in time.
In this formulation we begin at one end of a chromosome by
sampling a tree under the standard coalescent, and then reason
about how this tree is modified by recombination events
as we move along the genome. Although this algorithm is
significantly more complicated to simulate than Hudson's
backwards in time approach, it has led to substantial
insights into the process of recombination itself
and to the sequentially Markov Coalescent (SMC)
approximations~\citep{mcvean2005approximating,marjoram2006fast}
which are central to many modern methods.

@@

\section*{Evaluation of sampling probabilities}

The conditional sampling probability $p(G | D)$ of an ARG $G$ given an observed
realization of genetic diversity $D$ at its sampled leaves is a central concept in
model fitting and selection. Under the Bayesian paradigm, it defines the posterior
distribution $p(G | D) \propto p(D | G) p(G)$, where $p(D | G)$ is the likelihood
of the data $D$ given $G$, and $p(G)$ is a prior distribution for the underlying
ARG $G$. Under the frequentist paradigm, $p(G)$ can be interpreted as a regularizer,
and $p(G | D)$ as a regularized likelihood. In both cases, $p(G)$ is typically specified
as the sampling distribution of a stochastic process, such as the coalescent with
recombination. Essentially all of the model-based inference methods surveyed below
rely on evaluation of a likelihood and a prior. Two contemporary examples of
likelihood-based methods are  \cite{mahmoudi2022bayesian, guo2022recombination}.

In essentially all neutral cases in practical use, the likelihood $p(D | G)$ is the law of
a Poisson process on the ancestral material on the edges of the ARG, which implies
that mutation processes at two sites are conditionally independent given the
underlying ancestral trees at those sites. A prototypical example is 
\citet[Eq.\ (2)]{mahmoudi2022bayesian}. Hence, $p(D | G)$ can be evaluated
regardless of the precise data structure used to store $G$, provided that it encodes
at least all marginal ancestral trees along the sampled sequences.  Choosing the
right data structure can still have a dramatic effect on the efficiency with which
$p(D | G)$ can be evaluated \citep{mahmoudi2022bayesian}.

In contrast, evaluating $p(G)$ under the coalescent with recombination
(or related models) requires knowledge of the number of extant lineages and
the number of links available for recombination (i.e.\ ones at which a recombination
would split ancestral material) at each time along the ARG
\citep[Eq.\ (3)]{mahmoudi2022bayesian}. These cannot be uniquely determined
from just the marginal trees at each site because branches at multiple trees can be
contained in the same ancestral lineage. Instead, a richer data structure akin to the
EARG or GARG is needed. Moreover, because EARG does not explicitly store the flow
of ancestral material along its edges---a necessary piece of information for
computing the number of effective recombination links at each time---evaluating
$p(G)$ based on an EARG essentially requires turning it into a GARG at runtime.

The number of extant lineages at a given time is not explicitly represented even
in a GARG, and needs to be calculated from the node and edge tables.
For example, any edges with the same child and parent nodes are contained within
the same lineage. Under the coalescent with recombination, the two edges
originating from a recombinant child are also contained within a single lineage
until the time of their parent nodes, at which point they split. Nodes S, T, and U
in panel (a) of Figure \ref{fig-ancestry-resolution} provide a concrete example
of the latter: the edges from S to T and S to U are distinct, but correspond to only
a single lineage in the time interval they span. We emphasize that these
conventions grouping edges into lineages are not part of the GARG data structure.
While they are convenient for analysis using the coalescent with recombination,
other models, such as discrete time Wright--Fisher models with finite populations,
may benefit from different ways to associate edges with lineages.

@@

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/inference.pdf}
\end{center}
\caption{\label{fig-inferred-args}
ARGs inferred by different methods from a small dataset. Notes about these
ARGs.
}
\end{figure}
